---
title: "Comparisons"
---

```{r include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(chorddiag)
library(htmlwidgets)
library(igraph)
library(readr)
library(tidygraph)
library(tidyverse)

```

## Comparing Past vs. Present Conditions

Using LANDFIRE’s BpS products, we explore two different ways to visualize past vs. current vegetation patterns.

-   First, we present **changes in broad ecosystem types** using an interactive comparison diagram. To present these broad ecosystem trends more effectively, we combined classifications of Existing Vegetation Types (EVT) into broader categories. The Developed EVTs of high, medium, and low intensity, and Quarries-Strip Mines-Gravel Pits EVTs were combined into one "Developed" EVT. Exotic Herbaceous and Exotic Tree-shrub EVTs were combined into one "Exotics" EVT.

-   Second, we compare **amounts of succession classes** (past and present) for the most prevalent ecosystems.

## Comparing Broad Vegetation Trends


### Chord Diagram

In the chord diagram below, you can hover over an outer band (e.g., near the phrase 'Past Hardwood' on the right side of the chord diagram) to see how much of a historical area is still mapped as that type, and how much is different. 

For example, if you hover over the outer green "Past Hardwood-Conifer" band you will see that roughly half has transitioned to "Hardwood" and a substantial portion has transitioned to "Conifer" in current mapping. 

**Some trends you might notice:**

* Some of the 'past conifer' has transitioned to 'hardwoods'.
* The area of interest does not have much 'Developed' area mapped.  


**PAST** &emsp; &emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp; &emsp; &emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp; &emsp; &emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp; &emsp;&emsp; &emsp; &emsp;&emsp; &emsp;**PRESENT**
```{r chord, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, fig.width=6}
# read in data
chord_df <- read_csv("data/bps2evt_chord.csv")
#view(histFireGVchord)

#convert to matrix
matrix_df <- as.matrix(as_adjacency_matrix(as_tbl_graph(chord_df),attr = "ACRES"))

#clean up matrix (could be cleaner!)
matrix_df = subset(matrix_df, select = -c(1:9))

matrix_df <- matrix_df[-c(9:14),]

#make a custom color pallet #eb4034 (redish) #b0af9e(grey)

# ORIGINAL
groupColors <- c("#f5e942", # current ag
                 "#1d4220", # current coni
                 "#397d3f", # current con-hard
                 "#b0af9e", # current deve
                 "#eb4034", # current exotics 
                 "#fc9d03", # current grass
                 "#56bf5f", # current hard
                 "#7db7c7", # current riparian

                
                 "#1d4220", # past coni
                 "#fc9d03", # past grass 
                 "#56bf5f", # past hard
                 "#397d3f", # past con-hard
                 "#7db7c7" # past riparian
                

)





#make chord diagram
chord <- chorddiag(data = matrix_df,
                 type = "bipartite",
                 groupColors = groupColors,
                 groupnamePadding = 10,
                 groupPadding = 3,
                 groupnameFontsize = 12 ,
                 showTicks = FALSE,
                 margin = 145,
                 tooltipGroupConnector = "    &#x25B6;    ",
                 chordedgeColor = "#363533"
                 )
chord 

#save then print to have white background
htmlwidgets::saveWidget(chord,
                        "chord.html",
                        background = "white",
                        selfcontained = TRUE
                        )
```



<iframe src="chord.html" height="720" width="720" style="border: 1px solid #464646;" allowfullscreen="" allow="autoplay" data-external=".5"></iframe>

<br>

## Succession classes for three most dominant BpSs

### Summary

* Laurentian-Acadian Northern Hardwoods Forest - Hemlock has seen a total loss of succession class D the tallest, most closed canopy class), largely replaced by class E, which is a mid-development class with species such as sugar maple (Acer saccharum), American basswood (Tilia americana), and paper birch (Betula papyrifera).
* Laurentian-Acadian Northern Pine(-Oak) Forest currently has an under representation of classes C and D (mid-open and late-closed classes), with an over representation of the B class (mid-closed).


```{r scls chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}

BPS_SCLS2 <- read.csv("data/bpsScls2.csv")

top_3_groups <- BPS_SCLS2 %>%
  group_by(BpS_Name) %>%
  summarize(total_count = sum(Freq, na.rm = TRUE)) %>%
  top_n(3, total_count) %>%
  pull(BpS_Name)

bps_scls_3 <- BPS_SCLS2 %>%
  group_by(Var1) %>%
  #mutate(total.count = sum(Freq)) %>%
  ungroup() %>%
  filter(BpS_Name %in% top_3_groups) %>%
  dplyr::select(c("BpS_Name", "refLabel",  "currentPercent", "refPercent")) %>%
  pivot_longer(
    cols = c(`refPercent`, `currentPercent`),
    names_to = "refCur",
    values_to = "Percent"
    )

write.csv(bps_scls_3, "final_df.csv")

# order classes
bps_scls_3$refLabel <- factor(bps_scls_3$refLabel, levels= c(
  "Developed",
  "Agriculture",
  "UE",
  "UN",
  "E",
  "D",
  "C",
  "B",
  "A"))




sclasplot <-
  ggplot(bps_scls_3, aes(fill=factor(refCur), y=Percent, x=refLabel)) +
  geom_col(width = 0.8, position = position_dodge()) +
  coord_flip() +
  facet_grid(. ~BpS) +
  scale_x_discrete(limits = (levels(bps_scls_3$refLabel))) +
  labs(
    title = "Succession Classes past and present",
    subtitle = "3 BpSs selected for illustration. Not all succession classes present in all BpSs.",
    caption = "Data from landfire.gov; Chart © Randy Swaty",
    x = "",
    y = "Percent")+
  theme_minimal(base_size = 14)+
  theme(plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") +
  scale_fill_manual(values = c("#3d4740", "#32a852" ), # present (grey), historical (green)
                    name = " ",
                    labels = c("Present",
                               "Past")) +
  facet_wrap(~BpS_Name, nrow(3),labeller = labeller(BpS_Name = label_wrap_gen())) +
    theme(panel.spacing = unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_rect(color = "black", size = 1))

sclasplot

```
